---
title: 'Updated combined annotation with CAMERA2 patient-episode analysis'
author: "Stefano Giulieri"
date: "03/05/2022"
output: 
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd(stringr::str_c(here::here(), "/eLife_revision/Updated_analysis_with_PROVEAN"))
msg <- glue::glue("My directory is {getwd()}")
message(msg)
```


Here we generate figure for the gene enrichment analysis. The whole gene convergence analysis process is described in `Gene_convergence_analysis.Rmd`.

```{r}
library(tidyverse)
library(patchwork)
library(ggrepel)
rm(list = ls())
```

# Load data

```{r}
df_gsea_neb <- readRDS("processed_data/gene_convergence/df_gsea_neb.Rda")

df_mutations <- readRDS("processed_data/gene_convergence/df_changes_multi_annot_for_gene_converg.Rda")
```

# Figure 3: gene enrichment in the entire dataset

```{r}
base_sz <- 20
point_sz <- 3
colors <- RColorBrewer::brewer.pal(n = 8, name = "PuOr")
names(colors) <- rev(c("IS insertion (intragenic)", "IS insertion (intergenic)", "large deletion", "intergenic", NA, "non-synonymous (neutral)", "non-synonymous (deleterious)", "truncating"))
colors_genes <- colors[c(1,2,3,8)]
colors_genes

df_gsea_all <- df_gsea_neb %>%
  filter(mutation_compartment == "all")
```

We want to add the fraction of deleterious mutations

```{r}
df_delet_frac <- df_mutations %>%
 mutate(deleterious_mutation = fct_collapse(EFFTYPE_SHORT_2, 
                                            deleterious = c("non-synonymous (deleterious)", "truncating", "IS insertion (intragenic)"),
                                            other_level = "non deleterious")) %>%
  distinct(mutation_id, neb_gene_symbol, deleterious_mutation) %>%
  dplyr::count(neb_gene_symbol, deleterious_mutation) %>%
  tidyr::complete(neb_gene_symbol, deleterious_mutation, fill = list(n = 0)) %>%
  dplyr::group_by(neb_gene_symbol) %>%
  mutate(frac = n/sum(n)) %>%
  filter(deleterious_mutation == "deleterious") %>%
  rename(frac_deleterious = frac)

df_mean_provean <- df_mutations %>%
  distinct(mutation_id, neb_gene_symbol, provean_score) %>%
  dplyr::group_by(neb_gene_symbol) %>%
  summarise(mean_provean_score = mean(provean_score, na.rm = T),
            median_provean_score = median(provean_score, na.rm = T))

df_gsea_all <- df_gsea_all %>%
  left_join(df_delet_frac) %>%
  left_join(df_mean_provean)

rm(df_delet_frac, df_mean_provean)
```


## Panel A 

```{r}
bonf <- .05/n_distinct(df_gsea_all$neb_locus_tag)
n_genes_with_sub <- df_gsea_all %>%
  filter(nsub > 0) %>%
  .$neb_locus_tag %>%
  n_distinct()
bonf <- .05/n_genes_with_sub
suggest <- 1/n_genes_with_sub

df_plot <- df_gsea_all %>%
  filter(enrich > 1) %>%
  select(neb_gene_symbol, glm.pval, enrich, frac_deleterious, mean_provean_score, median_provean_score) %>%
  arrange(glm.pval) %>%
  mutate(glm.pval = -log10(glm.pval),
         plot_order = row_number()) 

df_plot <- df_plot %>%
  filter(plot_order <= 20)

p1 <- df_plot %>%
  ggplot(aes(x = fct_reorder(neb_gene_symbol, plot_order, .desc = T),
             y = glm.pval, fill = frac_deleterious)) +
  geom_point(aes(size = enrich), shape = 21) +
  geom_hline(yintercept = -log10(bonf), linetype = "dashed") +
  scale_size(name = "Enrichment", breaks = c(10, 20, 40)) +
  colorspace::scale_fill_continuous_diverging("Blue-Red 3", mid = .5, name = "Proportion deleterious",
                                              breaks = c(0,.5,1)) +
  coord_flip() +
  labs(x = "",
       y = "-log10(p)") +
  theme_bw(base_size = base_sz) +
  theme(text = element_text(face = "bold"),
        legend.position = "right",
        legend.title = element_text(size = 16))
p1
```

# Panel B

```{r}
df_plot2 <- df_mutations %>%
  group_by(neb_gene_symbol, EFFTYPE_SHORT_2, mutation_compartment) %>%
  summarise(n_mutations = n_distinct(patient_episode_unique))

df_plot2 <- df_plot2 %>%
  right_join(df_plot %>% select(neb_gene_symbol, plot_order))

# change the label of the mutation_compartment
df_plot2 <- df_plot2 %>%
  ungroup() %>%
  mutate(mutation_compartment = fct_recode(mutation_compartment,
                                           `type CC` = "colonising-colonising",
                                           `type CI` = "colonising-invasive",
                                           `type II` = "invasive-invasive"))

p3 <- df_plot2 %>%
  ggplot(aes(x = fct_reorder(neb_gene_symbol, plot_order, .desc = T),
             y = n_mutations,
             fill = EFFTYPE_SHORT_2)) +
  geom_bar(stat = "identity") +
  facet_wrap(~mutation_compartment) +
 scale_fill_manual(values = colors_genes, name = "Mutation", labels = ~str_remove(., "onymous")) +
  scale_y_continuous(breaks = seq(0,15, 5),
                     minor_breaks = seq(0,15, 1)) +
  labs(x = "", y = "Independent acquisitions") +
  coord_flip() +
  theme_bw(base_size = base_sz) +
  theme(text = element_text(face = "bold"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.title = element_text(size = 16))

p3
```

Plot B: mutation mini-maps

Here we plot the whole operon with location of intragenic and intergenic mutations. For this, we need to extract the location of intergenic mutations in relation to the flanking protein. We already have this information for point mutations but not for large deletions and IS insertions.

```{r}
df_gene_maps <- df_mutations %>%
  right_join(df_plot %>% select(neb_gene_symbol, plot_order)) %>%
  group_by(neb_gene_symbol, AA_POS, EFFTYPE_SHORT_2, plot_order) %>%
  summarise(n_events = n_distinct(patient_episode_unique)) 

df_rect <- df_mutations %>%
  right_join(df_plot %>% select(neb_gene_symbol, plot_order)) %>%
  group_by(neb_gene_symbol, plot_order) %>%
  summarise(xmin = plot_order - 0.25,
            xmax = plot_order + 0.25, 
            ymin = 0,
            ymax = max(GENE_LENGTH))
  

p4 <- df_gene_maps %>%
  ggplot(aes(x = plot_order, y = AA_POS, colour = EFFTYPE_SHORT_2, size = n_events)) +
  geom_point(shape = 17) +
  geom_rect(data = df_rect, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = F, alpha = 0, color = "black") +
  scale_y_continuous(limits = c(0,1000), breaks = c(0,500, 1000)) +
  scale_x_reverse(expand = c(0, .4)) +
  scale_color_manual(values = colors_genes, na.translate = F, name = "Mutation", guide = F) +
  scale_size(breaks = 1:3, name = "Independent\nacquisitions") +
  coord_flip() +
  labs(x = "",
       y = "Amino acid position") +
  theme_bw(base_size = base_sz) +
  theme(panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.position = "right") +
  theme(text = element_text(face = "bold"),
        legend.title = element_text(size = 16))

p4
```

Combine

```{r}
wrap_plots(p1, p3, p4) +
  plot_annotation(tag_levels = "A") +
  plot_layout(widths = c(5, 15, 11),  guides = "collect") &
  theme(plot.tag = element_text(size = 20))

dir.create("figures/")
ggsave("figures/gene_enrichment_all.pdf", width = 15, height = 7)

# Save the plot objects
saveRDS(p1, "processed_data/gene_convergence/gene_convergence_p1.Rda")
```

# Supplementary figure: deleterious mutations in convergent genes

```{r}
df_plot3 <- df_plot2 %>%
  dplyr::group_by(neb_gene_symbol, EFFTYPE_SHORT_2) %>%
  mutate(EFFTYPE_SHORT_2 = fct_collapse(EFFTYPE_SHORT_2, truncating = c("truncating", "IS insertion (intragenic)"))) %>%
  summarise(n_mutations = sum(n_mutations)) %>%
  dplyr::group_by(neb_gene_symbol) %>%
  mutate(frac = n_mutations/sum(n_mutations),
         total_mutations = sum(n_mutations)) %>%
  complete(neb_gene_symbol, total_mutations, EFFTYPE_SHORT_2, fill = list(frac = 0) ) %>%
  filter(EFFTYPE_SHORT_2 == "truncating") %>%
  rename(frac_truncating = frac) %>%
  left_join(df_plot)

pos <- position_jitter(seed = 2)
df_plot3 %>%
  ggplot(aes(x = frac_deleterious, y = frac_truncating)) +
  geom_jitter(aes(size = total_mutations, fill = median_provean_score), shape = 21, position = pos) +
  ggrepel::geom_text_repel(aes(label = neb_gene_symbol), box.padding = .25, position = pos, size = 5, fontface = "bold") +
  scale_size(name = "Number of mutations", range = c(1,10)) +
  scale_fill_gradient2(midpoint = -2.5, name = "Median PROVEAN score") +
  labs(x = "Proportion deleterious", y = "Proportion truncating") +
  theme_bw(base_size = 16) +
  theme(text = element_text(face = "bold"))

ggsave("figures/gene_delet_trunc_scatter_plot.pdf", width = 10, height = 7)
```

Summary of these data

```{r}
df_mutations %>%
  inner_join(df_plot) %>%
  distinct(mutation_id, EFFTYPE_SHORT_2, neb_gene_symbol) %>%
  mutate(deleterious_mutation = fct_collapse(EFFTYPE_SHORT_2, 
                                            deleterious = c("non-synonymous (deleterious)", "truncating", "IS insertion (intragenic)"),
                                            other_level = "non deleterious")) %>%
  dplyr::count(deleterious_mutation) %>%
  ungroup() %>%
  mutate(frac = n/sum(n))
```

This figure could be further improved by chosing the genes listed in table 2 and 3. This would, for example, allow to include *rpoB* which has a very interesting profile: no truncations and high proportion of deleterious substitutions.

