---
title: 'Updated operon convergence analysis with CAMERA2 patient-episode analysis: create figures'
author: "Stefano Giulieri"
date: "11/10/2021"
output: 
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd(stringr::str_c(here::here(), "/Updated_analysis_with_CAMERA2"))
msg <- glue::glue("My directory is {getwd()}")
message(msg)
```


Here we generate figure for the gene enrichment analysis. The whole gene convergence analysis process is described in `Operon_convergence_analysis.Rmd`.

```{r}
library(tidyverse)
library(patchwork)
library(ggrepel)
rm(list = ls())
```

# Load data

```{r}
df_gsea_neb <- readRDS("processed_data/operon_convergence/df_gsea_neb.Rda")

df_mutations <- readRDS("processed_data/operon_convergence/df_changes_multi_annot_for_operon_converg.Rda")
```

# Figure 3: gene enrichment in the entire dataset

```{r}
base_sz <- 20
point_sz <- 3
colors <- RColorBrewer::brewer.pal(n = 6, name = "PuOr")
names(colors) <- rev(c("IS insertion (intragenic)", "IS insertion (intergenic)", "large deletion", "intergenic", "non-synonymous", "truncating"))
colors_genes <- colors[c(1,2,3,5,6)]
colors_genes

df_gsea_all <- df_gsea_neb %>%
  filter(mutation_compartment == "all")
```


## Panel A 

```{r}
bonf <- .05/n_distinct(df_gsea_all$neb_operon_symbol)
suggest <- bonf*20
n_operons_with_sub <- df_gsea_all %>%
  filter(nsub > 0) %>%
  .$neb_operon_symbol %>%
  n_distinct()
# bonf <- .05/n_genes_with_sub
# suggest <- 1/n_genes_with_sub

df_plot <- df_gsea_all %>%
  filter(enrich > 1) %>%
  distinct(neb_operon_symbol, glm.pval, enrich) %>%
  arrange(glm.pval) %>%
  mutate(glm.pval = -log10(glm.pval),
         plot_order = row_number())

df_plot <- df_plot %>%
 filter(plot_order < 20)

# Label with the number of mutations. Limit the maximum number of genes so to avoid huge labels
df <- df_gsea_all %>%
  mutate(operon_label_complete = str_c(str_trunc(neb_operon_symbol, width = 20), " (", nsub, ")" )) %>%
  distinct(neb_operon_symbol, operon_label_complete)



# operon_label_complete <- str_wrap(df$operon_label_complete, width = 40)
operon_label_complete <- df$operon_label_complete
names(operon_label_complete) <- df$neb_operon_symbol

  
  
p1 <- df_plot %>%
  ggplot(aes(x = fct_reorder(neb_operon_symbol, plot_order, .desc = T),
             y = glm.pval, colour = glm.pval > -log10(bonf))) +
 # geom_point(size = point_sz) +
  geom_point(aes(size = enrich)) +
  geom_hline(yintercept = -log10(bonf), linetype = "dashed") +
  # geom_hline(yintercept = -log10(suggest), colour = "red") +
  scale_size(name = "Enrichment", breaks = c(5, 10)) +
  scale_colour_manual(values = c("blue", "red"), labels = c("> Bonf", "< Bonf"), name = "p value", guide = "none") +
  scale_x_discrete(labels = operon_label_complete) +
  scale_y_continuous(limits = c(NA,22.5), breaks = c(10,20)) +
  coord_flip() +
  labs(x = "",
       y = "-log10(p)") +
  theme_bw(base_size = base_sz) +
  theme(text = element_text(face = "bold"),
        legend.position = "right",
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 15),
        axis.text.y = element_text(size = 15))
p1
```

# Panel B

```{r}
df_plot2 <- df_mutations %>%
  group_by(neb_operon_symbol, EFFTYPE_SHORT, mutation_compartment) %>%
  summarise(n_mutations = n_distinct(patient_episode_unique))

df_plot2 <- df_plot2 %>%
  right_join(df_plot %>% select(neb_operon_symbol, plot_order))




# change the label of the mutation_compartment
df_plot2 <- df_plot2 %>%
  ungroup() %>%
  mutate(mutation_compartment = fct_recode(mutation_compartment,
                                           `type CC` = "colonising-colonising",
                                           `type CI` = "colonising-invasive",
                                           `type II` = "invasive-invasive"))

p3 <- df_plot2 %>%
  ggplot(aes(x = fct_reorder(neb_operon_symbol, plot_order, .desc = T),
             y = n_mutations,
             fill = fct_rev(fct_relevel(EFFTYPE_SHORT, names(colors_genes))))) +
  geom_bar(stat = "identity") +
  facet_wrap(~mutation_compartment) +
 scale_fill_manual(values = colors_genes, name = "Mutation", labels = ~str_remove(., " insertion")) +
  scale_y_continuous(limits = c(0,15), breaks = c(0,10,20), minor_breaks = scales::breaks_width(2)) +
  scale_x_discrete(labels = operon_label_complete) +
  labs(x = "", y = "N independent mutations") +
  coord_flip() +
  theme_bw(base_size = base_sz) +
  theme(text = element_text(face = "bold"),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 16))

p3


# p1 + p3 +
#   plot_layout(guides = "collect")
```


Plot B: mutation mini-maps

Here we plot the whole operon with location of intragenic and intergenic mutations. For this, we need to extract the location of intergenic mutations in relation to the flanking protein. We already have this information for point mutations but not for large deletions and IS insertions.

```{r}
df_rect <- df_gsea_all %>%
  select(neb_operon_symbol, neb_operon_start, neb_operon_end, neb_operon_strand, neb_operon_gene_position, ymin = neb_operon_gene_start, ymax = neb_operon_gene_end,neb_locus_tag, neb_gene_symbol) %>%
  right_join(df_plot %>% select(neb_operon_symbol, plot_order)) %>%
  mutate(xmin = plot_order - 0.25,
            xmax = plot_order + 0.25) 

# df_plot3 <- df_mutations %>%
#   left_join(df_rect) %>%
#   mutate(y = if_else(mutation_position_to_flanking_prot == "promoter", distance_from_flanking_prot + ymin, distance_from_flanking_prot + ymax )) %>%
#   mutate(y = if_else(is.na(y), NT_POS + ymin, y)) %>%
#   relocate(neb_start, y, ymin, ymax)

df_operon_maps <- df_mutations %>%
  rename(y = operon_position) %>%
   right_join(df_plot %>% select(neb_operon_symbol, plot_order)) %>%
  group_by(neb_operon_symbol, y, EFFTYPE_SHORT, plot_order) %>%
  summarise(n_events = n_distinct(patient_episode_unique)) 

p4 <- df_operon_maps %>%
  ggplot(aes(x = plot_order, y = y, colour = EFFTYPE_SHORT, size = n_events)) +
  geom_point(shape = 17) +
  geom_rect(data = df_rect, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = F, alpha = 0, color = "black") +
  scale_x_reverse(expand = c(0, .4)) +
  scale_y_continuous(labels = ~ ./1000) +
  scale_color_manual(values = colors_genes, na.translate = F, guide = F) +
 scale_size(breaks = 1:3, name = "Independent\nacquisitions") +
  coord_flip() +
  labs(x = "", y = "Nucleotide position (kb)") +
  theme_bw(base_size = base_sz) +
  theme(panel.grid = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        text = element_text(face = "bold"),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 16))

p4
```

Combine

```{r}
wrap_plots(p1, p3, p4) +
  plot_annotation(tag_levels = "A") +
  plot_layout(widths = c(5, 15, 11),  guides = "collect") &
  theme(plot.tag = element_text(size = 20))

dir.create("figures/")
ggsave("figures/operon_enrichment_all.pdf", width = 15, height = 7)
```

# Figure 5: GSEA across compartments

```{r}
df_gsea_compartments <- df_gsea_neb %>%
  filter(mutation_compartment != "all")

# bonf <- bonf/3
# suggest <- suggest/3

# df_gsea_compartments <- df_gsea_compartments %>%
#   group_by(neb_locus_tag) %>%
#   filter(any(nsub > 0)) %>%
#   group_by(mutation_compartment) %>%
#   mutate(p_fdr = p.adjust(glm.pval, method = "fdr")) %>%
#   relocate(p_fdr)
```

As Volcano plot

```{r}
df_plot <- df_gsea_compartments %>%
  group_by(neb_operon_symbol) %>%
  filter(any(nsub > 0)) %>%
  mutate(mutation_compartment = fct_recode(mutation_compartment,
                                           `type CC` = "colonising-colonising",
                                           `type CI` = "colonising-invasive",
                                           `type II` = "invasive-invasive")) %>%
  # remove genes with zero value
  filter(enrich > 1) %>%
  distinct(neb_operon_symbol, glm.pval, enrich, mutation_compartment, neb_operon_start)


df_plot %>%
  ggplot(aes(x = enrich, y = -log10(glm.pval))) +
  geom_hline(yintercept = -log10(bonf), linetype = "dashed") + 
   geom_hline(yintercept = -log10(suggest), linetype = "dotted") + 
  geom_point(aes(colour = glm.pval < bonf)) +
  geom_label_repel(data = . %>% filter(glm.pval < suggest), aes(label = neb_operon_symbol),
                  box.padding = .25, min.segment.length = 0.25, point.size =1, force_pull = 2, force = 2,  
                  max.overlaps = Inf) +
  scale_x_continuous(trans = "log10") +
  scale_color_manual(values = c("blue", "red"), guide = F) +
  # ylim(0,7) +
  facet_wrap(~mutation_compartment,  nrow = 1) +
  labs(x = "Enrichment", y = "-log10(p value)") +
  # ylim(0,10) +
  theme_bw(base_size = 20) +
  theme(text = element_text(face = "bold"))

ggsave("figures/operon_enrichment_volcano_plot_compart_only_no_scale_limits.pdf", width = 10, height = 5)
```

As Manthattan plot

```{r}
df_plot %>%
  mutate(signif_category = case_when(
    glm.pval < bonf ~ "bonf",
    glm.pval > bonf & glm.pval < suggest ~ "suggest",
    glm.pval > suggest ~ "ns"
  )) %>%
  ggplot(aes(x = neb_operon_start/1e6, y = -log10(glm.pval))) +
  geom_hline(yintercept = -log10(bonf), linetype = "dashed") + 
   geom_hline(yintercept = -log10(suggest), linetype = "dotted") + 
  geom_point(aes(colour = glm.pval < bonf, size = enrich)) +
  geom_label_repel(data = . %>% filter(glm.pval < suggest), aes(label = neb_operon_symbol),
                  box.padding = .25, min.segment.length = 0, point.size =1, nudge_x = -.25) +
  scale_color_manual(values = c("blue", "red"), guide = F) +
  scale_size(breaks = c(10,20,40)) +
  # ylim(0,7) +
  facet_wrap(~mutation_compartment,  nrow = 1) +
  labs(x = "Gene position (Mb)", y = "-log10(p value)") +
  # ylim(0,10) +
  theme_bw(base_size = 20) +
  theme(text = element_text(face = "bold"))

ggsave("figures/gene_enrichment_manhattan_plot_compart_only_no_scale_limits.pdf", width = 10, height = 5)
```

