---
title: 'Updated combined annotation with CAMERA2 patient-episode analysis'
author: "Stefano Giulieri"
date: "08/10/2021"
output: 
  html_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd(stringr::str_c(here::here(), "/Updated_analysis_with_CAMERA2"))
msg <- glue::glue("My directory is {getwd()}")
message(msg)
```


Here we create a dataset with convergence data. The whole gene convergence analysis process is described in `Gene_convergence_analysis.Rmd`.

The gene convergence dataset will be assembled using:

1. The gsea output
2. The neb annotation
3. (optional) the list of mutations and isolates with mutations


```{r}
library(tidyverse)
library(BioNet)
rm(list = ls())
```

# Parse GSEA results

```{r}
file.copy("~/Documents/Transfer_with_server/genes", "GSEA/", recursive = T, overwrite = F)

f <- list.files("GSEA/genes", recursive = T,  pattern = "results.gene", full.names = T)

gene_results <- map_df(f, function(x){
  class <- basename(dirname(x))
  df <- read_tsv(x) %>%
    mutate(mutation_compartment = class)
})

f <- list.files("GSEA/genes", recursive = T,  pattern = "results.ontology", full.names = T)

ontology_results <- map_df(f, function(x){
  class <- basename(dirname(x))
  df <- read_tsv(x) %>%
    mutate(mutation_compartment = class)
})
```

Combine in one dataframe

```{r}
df_gsea <- gene_results %>%
  select(neb_locus_tag = Locus, nsub, kb, enrich, mutation_compartment) %>%
  left_join(ontology_results %>% select(neb_locus_tag = Ontology, glm.pval, mutation_compartment)) %>%
  mutate(mutation_compartment = as.factor(mutation_compartment)) %>%
  arrange(glm.pval) 
```

# Set significance threshold

```{r}
bonf <- .05/n_distinct(df_gsea$neb_locus_tag)
suggest <- 1/n_distinct(df_gsea$neb_locus_tag)
# bum.mle <- fitBumModel(ontology_results$glm.pval, plot=TRUE)
# tau <- fdrThreshold(fdr=0.05, fb=bum.mle)
```

# Add neb annotation

Phage will be removed because we perform an `inner_join()`

```{r}
df_neb_curated <- readRDS("../../SAUREUS-GENERAL/ref_genomes/processed_data/FPR3757/df_neb_curated.Rda")

df_gsea_neb <- df_gsea %>%
  inner_join(df_neb_curated) %>%
  relocate(neb_locus_tag, neb_gene_symbol) 
```

# Add mutations

Here we need to perform a `left_join()`. Why? Because x (`df_gsea`) contains rows that don't exist in y (`df_mutations`), namely genes with no mutations and rows with a `all` compartment

```{r}
df_mutations <- readRDS("processed_data/gene_convergence/df_changes_multi_annot_for_gene_converg.Rda")

df_gsea_neb_with_mutations <- df_gsea_neb %>%
  left_join(df_mutations)
```

# Save processed data

```{r}
df_gsea_neb %>%
  saveRDS("processed_data/gene_convergence/df_gsea_neb.Rda")
df_gsea_neb %>%
  write_csv("processed_data/gene_convergence/df_gsea_neb.csv")

df_gsea_neb_with_mutations %>%
  saveRDS("processed_data/gene_convergence/df_gsea_neb_with_mutations.Rda")
df_gsea_neb_with_mutations %>%
  write_csv("processed_data/gene_convergence/df_gsea_neb_with_mutations.csv")
```

